Coalition Formation Algorithm of Prosumers in a Smart Grid Environment

### 2015-02-09 Article.tex
typo
minipage update
big updates

@@

One of the key ideas in the smart grid revolves around the introduction of communication means inside the power grids. This could enable complex improvements in the energy management and leads progressively to a greener energy system \cite{Ramchurn} \cite{WuHamedHuangBook2011}. Distributed energy resources (DER) such as wind turbines or photovoltaic panels are not supposed to emerge only in remote farms, but also in residential areas. Together with electric vehicles, and demand side management programs \cite{Samadi2014}, they will constitute the building blocks which will help to turn the today pure energy consumers into true actors of the grid operation \cite{Ramchurn}. Such agents that both consume and produce energy are ready to make concessions (appliances delays, V2G) to ensure grid stability, and are commonly called prosumers \cite{6883384,Ramchurn}.

@@

In this paper, we consider the case where prosumers wanted to be clustered together into coalitions in order to be able to produce enough power which may be sold to the grid. If the coalition's production is sufficiently high and stable, the coalition can negociate on a day by day basis with the grid by announcing each day an expected production for the next day.

@@

In this paper we do not consider stability against player defection but focus on a statistics-oriented definition of coalition stability. We want to form coalitions based on utilities that depend on statistical properties of time-series values. The setting is thus similar with some financial studies on stock exchanges, where researchers tried to find relevant correlated clusters of stocks based on their daily prices variations. There exists well-known algorithms such as K-means or hierarchical clustering that are traditionally used for such purposes. Nevertheless, despite their popularity, they do not appear particularly well suited for meaningfully translating complex correlation relationships between time series while clustering them, especially when these series exhibit complicated underlying patterns. We will show later in this paper that using K-means as a clustering technique for grouping prosumers together leads to less good results.

@@

\begin{center}
\begin{figure}
\includegraphics[scale=0.45]{figure2/Fig2}
\caption{Process diagram}
\label{Fig1}
\end{figure}
\end{center}

@@

In the following, we do not make any gaussian assumption as in the previous toy example. We investigate then the developement of a heuristic that tries to find a compromise between maximizing the production sold on the market (benefits) on one hand (see eq. \ref{problem_1}) and minimizing the probability that the coalitions fail to sustain their contract values (stability) on the other hand (see eq. \ref{problem_2}). Table \ref{table1} summarizes the main notations.

@@

\begin{table}[t]
\centering
\scriptsize
\begin{tabular}{l|p{4.5cm}}
  \hline 
  \\
  $ d_{ij} $ & Distance metric between timeseries i and j \\
  $ \rho_{ij} $ & Pearson's correlation coefficient between timeseries i and j \\
  $ \mathcal{A} $ & Agent set \\
  $ P_{i}(t) \in \mathbb{R} $ & Instantaneous production of agent $i$ at time $t$ \\
  $ P_{s}(t) $ & Instantaneous production of coalition s \\
  $ \mathcal{F}(P_{i}) $ & Probability distribution of $i$'s production \\
  $ \mathcal{F}(P_{S}) $ & Probability distribution of coalition s production \\
  $ \mathcal{N}(\mu, \sigma) $ & Normal distribution with mean $ \mu $ and standard deviation $ \sigma $ \\
  $ \phi \in [0,1] $ & Reliability threshold \\
  $ P^{MIN} \in \mathbb{R}^+ $ & Min value threshold \\
  $ P_{s}^{CRCT} \in \mathbb{R}^+ $ & Contract value announced by coalition s \\
  $ P_{\phi}(s) \in \mathbb{R}^+ $ & Max contract value under $ \phi $ constraint \\
  $ \mathcal{U}_{\phi,\ P^{MIN}}(s) $ & Utility function \\
  $ \tau_{i}(s) \in \mathbb{R}$ & Overlapping to disjoint coalitions mapping \\
  $ \epsilon \in [0,1] $ & Pruning threshold for the correlation graph \\
  $ G_{\epsilon}(\mathcal{A}, E) $ & Filtered correlation graph \\
  $ N_{COAL} \in \mathbb{Z}^+ $ & Number of desired coalition \\
  $ CS $ & Set of all coalition structures \\
  $ \Theta(G) $ & Set of non overlaping cliques in graph G \\
  $ \mathbf{1}_{X} $ & Indicator function over set X \\ 
  \\
  \hline
\end{tabular}

@@

\subsection{Utility function}\label{subsec:UtilityFunc}
Adressing the problem of eq.\ref{problem_1} and \ref{problem_2}, we choose a utility function (eq. \ref{utility}) that derives directly from the above remarks. If a coalition cannot provide a valid contract value, it receives naturally a utility of zero. Furthermore, the utility increases with the overall production (sum of the contracted values) as stated in the objective of eq.\ref{problem_1}. The stability objective described in eq.\ref{problem_2} is also present in this utility function through the $ \phi $ parameter. Finally, the $ 1/|s| $ term in eq. \ref{utility} indicates that we favor small coalitions, mainly because they are easier to maintain in terms of communications.

@@

Obviously, maximising this utility function amounts to maximizing the coalition contract value with the minimum possible number of agents. In such settings, and through the clique percolation procedure (section \ref{sec:forming}), we will show that coalitions with high utilities can be computed.

@@

However, these clusters of strongly correlated time-series are the exact opposite of what we are seeking. We can indeed consider them directly as coalitions and compute their utilities, and the results show (see the green curves with squares on figure \ref{Fig4}), as expected, terrible values. We will indeed consider in section \ref{sec:results} this case as a worst case scenario, far worse than a trivial random split of the agents in the same number of coalitions (see the red curves with triangles of the random case on figure \ref{Fig4}).

@@

However, as seems intuitively understandable, cliques of this graph tend to exhibit very good utility values. Such  structures contain indeed a link between every two nodes, meaning that the overall correlation is quite small.

Obviously, the $ \epsilon $ parameter is indirectly responsible for the sizes of the cliques : if it is too low, $ G_{\epsilon} $ does not provide enough cliques, conversely, if $\epsilon $ is too high, we loose important information as the graph becomes very dense and cliques tend to overlap strongly. For large values of $ \epsilon $ and for non trivial number of agents, finding cliques can even become computationally intensive.

Despite being direct and simple, improving the sizes of the cliques by increasing $ \epsilon $ does not seem elegant. Furthermore, as the utility function is also focused on maximizing $ P_{\phi}(s) $, it might be the case that some cliques benefit from additional agents even if they don't form a clique anymore. Cliques for small values of $ \epsilon $ appear thus as good seeds for forming stable coalitions rather than directly acceptable coalitions.

We define $ N_{COAL} $ as the desired number of coalitions fixed by the user. Furthermore, let $ \Theta(G_{\epsilon}) $ be the set of non overlaping cliques in $ G_{\epsilon} $, we choose the $ \epsilon $ parameter as the smallest value such that $ G_{\epsilon} $ contains at least $ N_{COAL} $ non overlaping cliques :

@@

We construct $ G_{\epsilon^{\star}} $ such that it contains at least $ N_{COAL} $ potential seeds that may or may not be above the grid requirements, but exhibit low correlations among the members. Naturally, grid requirements could eventually be reached and social welfare increased by incorporating more nodes into the seeds. For this purpose, we use a clique percolation algorithm on $ G_{\epsilon^{\star}} $ that will extend the seeds over a correlation constrained environment. The main idea behind this algorithm is to make the seeds grow by local optimization of a fitness function, usually based on internal and external degrees of the seeds \cite{Lancichinetti}. The main difference between our case study and this algorithm for community detection is that we chose the fitness function as our utility function (cf. eq.\eqref{utility}) that both maximizes the production and minimizes its volatility (see eq.\ref{problem_1} and eq.\ref{problem_2}).

@@

In more details, figure \ref{Fig4}a presents the evolution of social welfare as the number of coalitions increases (all other parameters are kept constant) for random (red curve with triangles), clique percolation (blue curve with circles), and correlated coalitions (green curves with squares) that stands for a worst case scenario.

As for figure \ref{Fig4}b, it shows the percentage of coalitions able to enter the market for different values of $ P^{MIN} $. For consistency, we average the results of both plots over 100 realizations (colored zones around the curves shows the standard deviations). When the grid requirements are constant (figure \ref{Fig4}b), and the number of desired coalitions is low, clique percolation generally performs better than random case. Moreover, when $ N_{COAL} $ gets bigger, the performance of a random split tumbles down rapidly while our clique percolation is able to maintain efficiently the social welfare of the coalitions formed. When the grid requirements vary, for very low $ P^{MIN} $, all coalitions for all algorithms are able to enter the market, yielding an acceptance percentage of $ 100 \% $. But as $ P^{MIN} $ increases, we see the percentage of the correlated coalitions is going down very quickly.

@@

\begin{figure}
 \centering
  \includegraphics[scale=0.36]{figure8/fig8_bis.pdf}
  \caption{Evolution of the social welfare \textbf{a)} and the percentage of coalitions able to enter the market \textbf{b)} for different values of $ P_{MIN} $. Blue curves with circles represent clique percolation, red curves with squares, the random process, and the green ones with triangles, correlated coalitions.}
 \label{Fig4}
\end{figure}

@@

\bibliographystyle{IEEEtran}  
\bibliography{Article}
\end{document}

>>>>

\usepackage{float}
\usepackage{multicol}

@@

One of the key ideas in the smart grid revolves around the introduction of communication means inside the power grids. This could enable complex improvements in the energy management and leads progressively to a greener energy system \cite{Ramchurn} \cite{WuHamedHuangBook2011}. Distributed energy resources (DER) such as wind turbines or photovoltaic panels will probably not only appear in remote farms, but also in residential areas. Together with electric vehicles, and demand side management programs \cite{Samadi2014}, they will constitute the building blocks which will help to turn the today pure energy consumers into true actors of the grid operation \cite{Ramchurn}. Such agents that both consume and produce energy are ready to make concessions (appliances delays, V2G) to ensure grid stability, and are commonly called prosumers \cite{6883384,Ramchurn}.

@@

In this paper, we consider the case where prosumers coalesce into virtual groups (the so-called coalitions) in order to be able to produce enough power which may be sold to the grid. If the coalition's production is sufficiently high and stable, the coalition can negociate on a day by day basis with the grid by announcing each day an expected production for the next day.

@@

In this paper we do not consider stability against player defection but focus on a statistics-oriented definition of coalition stability. We want to form coalitions based on utilities that depend on statistical properties of time-series values. The setting is thus similar with some financial studies on stock exchanges, where researchers tried to find relevant correlated clusters of stocks based on their daily prices variations. There exists well-known algorithms such as K-means or hierarchical clustering that are traditionally used for such purposes. Nevertheless, despite their popularity, they do not appear particularly well suited for meaningfully translating complex correlation relationships between time series while clustering them, especially when these series exhibit complicated underlying patterns. We will show later in this paper that using K-means as a clustering technique for grouping prosumers together leads to less good results (see fig. \ref{Fig4}).

@@

\begin{center}
\begin{figure}
\includegraphics[scale=0.51]{figure2/Fig2}
\caption{Process diagram}
\label{Fig1}
\end{figure}
\end{center}

@@

In the following, we do not make any gaussian assumption as in the previous toy example (see for instance the densities at the coalition level on figure \ref{Fig5}). We investigate then the developement of a heuristic that tries to find a compromise between maximizing the production sold on the market (benefits) on one hand (see eq. \ref{problem_1}) and minimizing the probability that the coalitions fail to sustain their contract values (stability) on the other hand (see eq. \ref{problem_2}). Table \ref{table1} summarizes the main notations.

@@

\begin{table}[t]
\centering
\scriptsize
\begin{tabular}{l|p{4.5cm}}
  \hline 
  $ d_{ij} $ & Distance metric between timeseries i and j \\
  $ \rho_{ij} $ & Pearson's correlation coefficient between timeseries i and j \\
  $ \mathcal{A} $ & Agent set \\
  $ P_{i}(t) \in \mathbb{R} $ & Instantaneous production of agent $i$ at time $t$ \\
  $ P_{s}(t) $ & Instantaneous production of coalition s \\
  $ \mathcal{F}(P_{i}) $ & Probability distribution of $i$'s production \\
  $ \mathcal{F}(P_{S}) $ & Probability distribution of coalition s production \\
  $ \mathcal{N}(\mu, \sigma) $ & Normal distribution with mean $ \mu $ and standard deviation $ \sigma $ \\
  $ \phi \in [0,1] $ & Reliability threshold \\
  $ P^{MIN} \in \mathbb{R}^+ $ & Min value threshold \\
  $ P_{s}^{CRCT} \in \mathbb{R}^+ $ & Contract value announced by coalition s \\
  $ P_{\phi}(s) \in \mathbb{R}^+ $ & Max contract value under $ \phi $ constraint \\
  $ \mathcal{U}_{\phi,\ P^{MIN}}(s) $ & Utility function \\
  $ \tau_{i}(s) \in \mathbb{R}$ & Overlapping to disjoint coalitions mapping \\
  $ \epsilon \in [0,1] $ & Pruning threshold for the correlation graph \\
  $ G_{\epsilon}(\mathcal{A}, E) $ & Filtered correlation graph \\
  $ N_{COAL} \in \mathbb{Z}^+ $ & Number of desired coalition \\
  $ CS $ & Set of all coalition structures \\
  $ \Theta(G) $ & Set of non overlaping cliques in graph G \\
  $ \mathbf{1}_{X} $ & Indicator function over set X \\ 
  \hline
\end{tabular}

@@

\subsection{Utility function}\label{subsec:UtilityFunc}
Adressing the problem of eq.\ref{problem_1} and \ref{problem_2}, we choose a utility function (eq. \ref{utility}) that derives directly from the above remarks. If a coalition cannot provide a valid contract value, it receives naturally a utility of zero. Furthermore, the utility increases with the overall production (sum of the contracted values) as stated in the objective of eq.\ref{problem_1}. The stability objective described in eq.\ref{problem_2} is also present in this utility function through the $ \phi $ parameter. Finally, the $ 1/|s| $ term in eq. \ref{utility} indicates that we favor small coalitions, mainly because they are easier to maintain.

@@

Obviously, maximizing this utility function amounts to maximizing the coalition contract value with the minimum possible number of agents. In such settings, and through the clique percolation procedure (section \ref{sec:forming}), we will show that coalitions with high utilities can be computed.

@@

However, these clusters of strongly correlated time-series are the exact opposite of what we are seeking. We can indeed consider them directly as coalitions and compute their utilities, and the results show (see the green curves with diamonds on figure \ref{Fig4}), as expected, terrible values. We will indeed consider in section \ref{sec:results} this case as a worst case scenario, far worse than a trivial random split of the agents in the same number of coalitions (see the red curves with triangles of the random case on figure \ref{Fig4}).

@@

However, as seems intuitively understandable, cliques of this graph tend to exhibit very good utility values. Such  structures contain indeed a link between every two nodes, meaning that the overall correlation is quite small. Obviously, the $ \epsilon $ parameter is indirectly responsible for the sizes of the cliques : if it is too low, $ G_{\epsilon} $ does not provide enough cliques, conversely, if $\epsilon $ is too high, we loose important information as the graph becomes very dense and cliques tend to overlap strongly. For large values of $ \epsilon $ and for non trivial number of agents, finding cliques can even become computationally intensive.

Despite being direct and simple, improving the sizes of the cliques by increasing $ \epsilon $ does not seem elegant. Furthermore, as the utility function is also focused on maximizing $ P_{\phi}(s) $, it might be the case that some cliques benefit from additional agents even if they don't form a clique anymore. Cliques for small values of $ \epsilon $ appear thus as good seeds for forming stable coalitions rather than directly acceptable coalitions. We define $ N_{COAL} $ as the desired number of coalitions fixed by the user. Furthermore, let $ \Theta(G_{\epsilon}) $ be the set of non overlaping cliques in $ G_{\epsilon} $, we choose the $ \epsilon $ parameter as the smallest value such that $ G_{\epsilon} $ contains at least $ N_{COAL} $ non overlaping cliques :

@@

We construct $ G_{\epsilon^{\star}} $ such that it contains at least $ N_{COAL} $ potential seeds that may or may not be above the grid requirements, but exhibit low correlations among the members. Naturally, grid requirements could eventually be reached and social welfare increased by incorporating more nodes into the seeds. For this purpose, we use a clique percolation algorithm on $ G_{\epsilon^{\star}} $ that will extend the seeds over a correlation constrained environment. The main idea behind this algorithm is to make the seeds grow by local optimization of a fitness function, usually based on internal and external degrees of the seeds \cite{Lancichinetti}. The main difference between our case study and this algorithm for community detection is that we chose the fitness function as our utility function (see eq.\eqref{utility}) that both maximizes the production and minimizes its volatility (see eq.\ref{problem_1} and eq.\ref{problem_2}).

@@

\begin{figure}
  \centering
  \includegraphics[scale=0.53]{./figure11/densities.pdf}
  \caption{Production probability densities for four coalitions. One obtained with our algorithm based on clique percolation (blue curve with circles), one with a random formation (red curve with triangles), another one with a K-means approach (yellow curve with squares), and finally one with correlated agents (green curves with diamonds).}
  \label{Fig5}
\end{figure}

In the following, we consider and compare four different methods : our method based on clique percolation (blue curve with circles), the previous random aggregation of agents (red curve with triangles), a K-means clustering approach (yellow curve with squares), and correlated coalitions (green curves with diamonds) obtained with a correlation graph and that stands for a "\textit{worst case scenario}".

As visible on figure \ref{Fig5}, the production densities at the coalition level exhibit very different shapes. While our algorithm results in a nice density with a low (compared to the other curves) standard deviation, the other methods seem to form coalitions that will have a very large spectrum for their production. In more details, figure \ref{Fig4}a presents the evolution of social welfare as the number of coalitions increases (all other parameters are kept constant) for the same four methods. As for figure \ref{Fig4}b, it shows the percentage of coalitions able to enter the market for different values of $ P^{MIN} $. For consistency, we average the results of both plots over 100 realizations (colored zones around the curves shows the standard deviations). When the grid requirements are constant (fig \ref{Fig4}b), and the number of desired coalitions is low, clique percolation performs better than random case. Moreover, when $ N_{COAL} $ gets bigger, the performance of a random split tumbles down rapidly while our clique percolation is able to maintain efficiently the social welfare of the coalitions formed.

\begin{figure}
\begin{minipage}[c]{\textwidth}
	\begin{center}
  		\includegraphics[scale=0.45]{figure10/perc2.pdf}
  			\caption{Evolution of the social welfare when the number of coalitions is varying \textbf{a)}, and the percentage of coalitions able to enter the market for different values of $ P_{MIN} $ \textbf{b)}. Blue curves with circles and red curves with triangles represent respectively the results for clique percolation and the random process. For comparison, the green curves with diamonds and the yellow ones with squares show respectively the results obtained with coalitions of correlated agents and with a K-means clustering.}
 		\label{Fig4}
	\end{center}
\end{minipage}

\end{figure}

\begin{minipage}{\textwidth}
\begin{multicols}{2}
When the grid requirements vary, for very low $ P^{MIN} $, all coalitions for all algorithms are able to enter the market, yielding an acceptance percentage of $ 100 \% $. But as $ P^{MIN} $ increases, we see the percentage of the correlated coalitions going down very quickly.

@@

@@

\bibliographystyle{IEEEtran}  
\bibliography{Article}

\end{multicols}

\end{minipage}

\end{document}
